### 2018-04-26 thesis_report/Thesis_Background.tex
Thesis report update.

\documentclass{llncs}
\usepackage[utf8]{inputenc}
\usepackage{todonotes}
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{dirtytalk}
\usepackage{color}
\usepackage{glossaries}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage[caption = false]{subfig}
\graphicspath{ {images/} }
\usepackage{rotating}
\usepackage{pdflscape}

% Reset footnotes every page
%\usepackage{perpage}
%\MakePerPage{footnote}

\begin{document}
\frontmatter

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%  HEADER  %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\title{Malware Detection Via Machine Learning}
\subtitle{}
\titlerunning{Malware Detection Via ML}
\author{João F. Godinho}
\institute{Instituto Superior Técnico\\
\email{joao.f.godinho@tecnico.ulisboa.pt}}
\let\oldaddcontentsline\addcontentsline
\def\addcontentsline#1#2#3{}
\maketitle
\def\addcontentsline#1#2#3{\oldaddcontentsline{#1}{#2}{#3}}
%\todo[inline]{Validate the title, subtitle}

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%% ACRONYMS %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\newacronym{hids}{HIDS}{Host-based Intrusion Detection Systems}
\newacronym{ml}{ML}{machine learning}
\newacronym{caro}{CARO}{Computer Antivirus Research Organization}
\newacronym{ai}{AI}{Artificial Intelligence}
\newacronym{api}{API}{Application Programming Interface}
\newacronym{bios}{BIOS}{Basic Input/Output System}
\newacronym{dll}{DLL}{Dynamic-Link Library}
\newacronym{cc}{CC}{Creative Common's}
\newacronym{ann}{ANN}{Artificial Neural Networks}
\newacronym{tpr}{TPR}{True Positive Rate}
\newacronym{fpr}{FPR}{False Positive Rate}
\newacronym{svm}{SVM}{Support Vector Machine}
\newacronym{mist}{MIST}{Malware Instruction Set}
\newacronym{lr}{LR}{Logistic Regression}
\newacronym{pe}{PE}{Portable Executable}

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%% ABSTRACT %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
	Malware, which focuses on harming or subverting a system, has grown exponentially in recent years. The regular signature and heuristic methods applied by anti-virus vendors can fall short on keeping up with the rise of new types of malware. Other works have shown that machine learning has potential for dealing with the task of malware detection. With this in mind, the current report elaborates on the problem of malware, how anti-virus detect it, and proposes the development of a machine learning system to deal with the task of malware detection. To design the system to be, the current work will make use of a dataset with 388,513 samples for training and evaluation, focusing on how laboratory and real-world conditions can affect the final result.
\end{abstract}

\setcounter{tocdepth}{2}
\tableofcontents

\mainmatter
%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%  INTRO   %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%

% Temporary quotes, move to bib
% 1 - Attacking Malicious Code: A report to the Infosec Research Council
% 2 - AV-TEST Security Report 2015/2016 - https://www.av-test.org/fileadmin/pdf/security_report/AV-TEST_Security_Report_2015-2016.pdf
% 3 - McAfee Labs Threats Report April 2017 - https://www.mcafee.com/us/resources/reports/rp-quarterly-threats-mar-2017.pdf
% 4 - Symantec Internet Security Threat Report - https://www.symantec.com/content/dam/symantec/docs/reports/istr-22-2017-en.pdf
% 5 - Scanners of The Year 2000: Heuristics - http://vxheaven.org/lib/adg00.html
% 6 - Thomas Mitchell - Machine Learning
% 7 - Some Studies in Machine Learning Using the Game of Checkers - Arthur L. Samuel - http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.368.2254&rep=rep1&type=pdf
% 8 - Prudent Practices for Designing Malware Experiments: Status Quo and Outlook

\section{Introduction}\label{sec:introduction}
Malware, which can be referred to by other names like malicious software or malicious code, can be described as \say{any code added, changed, or removed from a software system in order to
intentionally cause harm or subvert the intended function of the system}\cite{mcgraw:mal_code}. Malware can be a piece of code attached to a legitimate program, an independent program, or a combination of both. The reasoning that leads to malware creation varies greatly, some are created to demonstrate some vulnerability or concept and do not cause direct harm to systems, while others are used to steal information or render systems useless. The wide variety and quantity of now-a-day systems and the amount of sensitive information they store provide numerous opportunities to illegally make a profit out of subverting legitimate systems.

The amount of malware has shown an exponential increase since the early 2000s, from 1.7 million in 2005 to over 600 million by the end of 2016\cite{av-test:report,mcafee:report}. According to a 2007 study by McAfee\footnote{McAfee Study Finds 4 Percent of Search Results Malicious, by Frederick Lane, June 4, 2007 [http:\/\/www.newsfactor.com\/story.xhtml?story\_id=010000CEUEQO]} about 4\% of the query results by major search engines lead to potentially dangerous websites. A more recent study by Norton in 2010\footnote{Symantec Press Release, July 29, 2010 [https://www.symantec.com/en/aa/about/newsroom/press-releases/2010/symantec\_0729\_02]} showed that at least 10\% of top trending search terms returned malicious results. From the contact with malicious software, 1.1 billion identities (i.e. user's personal information) have been exposed in 2016 alone\cite{symantec:report}.

Given the wide range and quantity of systems being operated by naive users, in the sense that they lack the knowledge to distinguish malicious and legitimate software, the first line of defense against threats are \gls{hids}, commonly known as anti-virus solutions. Their main function is to prevent infections by malicious code, which is achieved by detecting a program as malicious and disallow it from running (i.e. quarantining the malicious file).

To detect malicious software, anti-virus solutions depend on two main methods. \textit{Signature-based} methods check for the presence of unique byte sequences from known malware. For this method to be effective, a database of all malware signatures must be available, which leads to an obvious shortcoming: only known malware is detected\cite{gryaznov:scanners}. New and unknown malware is not detected until vendors analyze it and update their clients. To improve the effectiveness of anti-virus and overcome the problem in aforementioned methods, \textit{heuristic-based} methods are employed. These apply a set of rules to decide whether a program is likely to be malware or not. Like before, a database of rules must be available, but unlike signature methods, which provide exact matching on malware, heuristic methods are capable, to some degree, of generalizing malware features, hence detecting new and unknown malware. Although more sophisticated, heuristic-based methods need a database of rules that must be defined \textit{a priori}, just like signature methods.

As mentioned before, signature methods only detect known malware, which translates to a very low false positive rate (i.e. goodware wrongly classified as malware), while having a high detection rate (i.e. correctly classified malware from the total malware), assuming the signature database is on par with malware samples, which at the rate new malware emerges becomes a hard task. For heuristic methods new and unknown malware can be detected due to similarities to known malware, translating to a higher detection rate with less information, but with the downside of raising the false positive rate, as benign programs (goodware) can be flagged malicious due to characteristics that fit into malware rules\cite{gryaznov:scanners}. To improve the effectiveness of anti-virus, both methods must complement each other.

Having the need to either create signatures for new malware or define a set of rules that describe malware characteristics becomes a cumbersome task for the increasing malware numbers. To try and improve the detection task of anti-virus software, the current work proposes a \gls{ml} approach to malware detection as an evolution to heuristic-based methods.

The machine learning field studies the development of computer programs that automatically improve with experience. A program is given a task and an experience, if its performance at the task improves with the experience given, then the program is learning the task\cite{mitchell:ml}. In the proposed work, the task at hand is to distinguish malware from goodware (i.e. classify samples) and the experience is a wide variety of samples, both goodware and malware. The final result is a program that performs well for the given task (i.e. successfully learns to distinguish goodware from malware).

The \gls{ml} field has been around for multiple decades, with some of the earlier works dating back to 1959 by Arthur L. Samuel\cite{samuel:studies}, but only in recent years has the field grown in popularity. Its recent growth can be accredited to two main factors: availability of large data sets and computation power. Larger data sets provide more experience for a program to learn from, but also require more processing power.

The growing number of malware samples provide a good opportunity to make use of \gls{ml} for malware detection. This approach is not new and has provided high detection rates with low false positive rates\cite{miller:rev_int,nissim:al_pdf,perdisci:behavior,rieck:dynamic,schultz:data_mining}. However, many of the resulting research is done under laboratory conditions that do not represent real world scenarios. Some of those conditions have been studied in \cite{rossow:practices} and include the correctness of datasets, transparency and realism. Alongside the application of \gls{ml} to detect malware, the proposed work will study how laboratory conditions differ from real world one's and how effectiveness is affected.

To provide a solution that makes use of \gls{ml} to detect malware and its performance under laboratory \textit{vs.} real world conditions, the proposed work will follow an approach that is divided into three stages:

\begin{itemize}
	\item \textit{Data collection}: experience from which a program can learn must be available, as such the first step must ensure the collection of samples that represent both malware and goodware examples. The current work will take advantage of the Malwr\cite{tool:malwr} service, which provides both static (e.g. library imports) and dynamic (e.g. library call order during runtime) details about a sample. The total number of available samples surpasses 388 thousand, dated from April 2013 to October 2016.
	\item \textit{Data labeling}: before training a classifier to detect malware, samples must be labeled as either benign or malign. The Malwr\cite{tool:malwr} service provides details on how multiple anti-virus solutions classify a sample, hence the approach is to use said classification to determine the legitimacy of a sample. This step is not linear, as different anti-virus classify samples differently, introducing edge cases (e.g. 50\% classify malware, the other 50\% classify goodware). To overcome this, the current work must develop a metric to deal with such cases.
	\item \textit{Classifier implementation}: at this stage the collected data is ready to be used as experience for the program. Using static and dynamic details together with the labeling of the samples, a classifier can be trained to distinguish between malware and goodware. Alongside the classifier creation, this phase must also measure the effectiveness of the solution. For this work's scope, the effectiveness is to be measured under laboratory and real world conditions.
\end{itemize}

In sum, the work to be done as to verify the application of \gls{ml} methods for the task of detecting malware is: collection of samples, labeling of collected samples and finally usage of the resulting dataset to achieve a malware classifier.

The following sections provide deeper analysis on the subject of using \gls{ml} to classify malware, as well as the choices to be made and the reasoning behind them. Section \ref{sec:related_work} describes related work that both influenced and motivated the current work. Section \ref{sec:solution} details the intended approach, the tools to be used, how the work is to be presented and the already accomplished work. Section \ref{sec:conclusion} summarizes and closes the report.

%\todo[inline]{falar dos antivirus para combater isto, falar nas tecnicas basicas: signature based, heuristic based. tirar partido de ML para estender metodos heuristicos e melhorar a classificaçao de malware.}
%
%
%\todo[inline]{one liner do que é malware (quote), quantificar (quote), necessidade de detectar malware mais rapidamente (automatizar/generalizar).
%
%Avanços na detecçao do malware, pouca coisa. Aplicaçao de ML neste tipo de problema.
%
%Contributos do trabalho para este problema, desenvolver um classificador que funcione no mundo real e ajude a resolver o problema.
%
%1. Dados recolhidos, o que posso extrair dos dados\\
%2. Definir o que é malware (thresholds, consenso)\\
%3. Comparaçao de metodologias (estatatica/dinamica)}

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%% RELATED  %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related Work}\label{sec:related_work}

% #1 Semantics-Aware Malware Detection
% #2 Attacking Malicious Code: A report to the Infosec Research Council
% #3 Malware, Rootkits & Botnets A Beginner's Guide pp.10
% #4 Computer viruses: theory and experiments
% #5 Theory of self-reproducing automata
% #6 The Evolution of Viruses and Worms
% #7 The art of computer virus research and defense
% #8 https://www.av-test.org/en/
% #9 http://www.caro.org/articles/naming.html
% #10 http://members.chello.at/erikajo/vnc99b2.txt
% #11 Learning to Detect and Classify Malicious Executables in the Wild
% #12 The Evolution of Viruses and Worms
% #13 http://cert.europa.eu/static/SecurityAdvisories/2017/CERT-EU-SA2017-012.pdf
% #14 https://www.virusbulletin.com/uploads/pdf/magazine/2002/200201.pdf

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Background
\subsection{Background}\label{subsec:background}
Malware, which can be referred to by other names like malicious software or malicious code, can be described as \say{a program that has malicious intent}\cite{christodorescu:semantics}, or more detailed as any code added, changed, or removed from a software system in order to intentionally cause harm or subvert the intended function of the system\cite{mcgraw:mal_code}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Malware detection
\subsection{Malware Detection}\label{subsec:mal_detec}

@@

\subsection{Cuckoo Sandbox and Malwr}\label{subsec:cuckoo}

@@

\begin{figure}[h]
	\includegraphics[width=\textwidth]{malwr_sample}
	\caption{Example of a behavioral report from Malwr}
	\label{fig:malwr_sample}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Machine Learning
\subsection{Machine Learning}\label{subsec:ml}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Machine Learning in malware detection
\subsection{Machine Learning in Malware Detection}\label{subsec:ml_md}

@@

Before analyzing the work done with \gls{ml} in malware detection, it is worth noting how such work is to be conducted. Rossow et al\cite{rossow:practices} present some guidelines for designing malware experiments. Their work surveys 36 academic publications from 2006 to 2011 that rely on malware execution. The authors note frequent shortcomings in the surveyed publications regarding assumptions on the use of execution-driven datasets, absence of security precautions taken during experiments and insufficient description of the experimental setup. The provided guidelines are split into four groups: correct datasets, transparency, realism and safety.

@@

\begin{figure}[h]
	\includegraphics[width=\textwidth]{mist}
	\caption{\gls{mist} representation as seen in \cite{rieck:dynamic}}
	\label{fig:mist}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%% SOLUTION %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\section{Solution}\label{sec:solution}

With the knowledge given from previous sections, the current section will describe in detail the work be done as to achieve the objective of detecting malware samples by applying \gls{ml} classifiers.

As mentioned in section \ref{sec:introduction}, the approach to be taken to achieve the desired system is divided into three states: data collection, data labeling and classifier implementation. The following will describe each stage with more detail, together with work that as already been done.

\subsection{Data Collection}\label{subsec:data_collection}

The core objective of this stage is to obtain a corpus from which information about malware can be extracted. As mentioned in subsection \ref{subsec:cuckoo}, reports from Malwr are to be used, as they provide both static and dynamic information. The choice of using the already available reports instead of generating reports from collected samples is two-fold, it saves the overhead time of doing analysis and removes any bias regarding the sample choice, as real-world submissions are being used. The downside of using these reports is assuming the analysis was correctly done, which may not hold and can introduce noise to the extracted information.

The data provided by each report can be either static or dynamic, as mentioned in subsection \ref{subsec:cuckoo}. Malwr also provides additional information regarding static analysis, as it checks if the sample was submitted to VirusTotal, if so it lists the classification given by multiple anti-virus vendors. This information can then be used to assert if a sample is malicious or not.

Although Malwr enables the user to submit various type of objects, this work will only deal with \gls{pe} files (e.g. 32bit executables, \gls{dll}'s), as the Cuckoo implementation used by Malwr only provides static analysis for such files.

To check if the \gls{pe} provided a good amount of samples, an initial scrape of the list of analysis\footnote{Malwr Analysis List [https://malwr.com/analysis/]} was made using Scrapy\cite{tool:scrapy} and the present file types were analyzed. At the time of scrape there was a total of 642,698 submissions, where 388,702 (60.48\%) were \gls{pe} files, dated between April 16th, 2013 and October 10th, 2016. Figure \ref{fig:dist_submissions} and \ref{fig:dist_submissions_types} show the submissions by month, the percentage of duplicates per month and how the sample types are distributed, worth noting is that in August 2014 no samples were collected, possibly due to service downtime.

Knowing that \gls{pe} still account for a good part of all samples, a scrape of those reports was made. The reports were saved as HTML in the gzip format to save disk space. A total of 388,513 reports were obtained (different from the previous amount as some errors could have occurred during scraping) occupying a total of 26GB.

Having the reports locally, the next step is to parse the HTML format to extract the information to be used. Related work\cite{miller:rev_int,nissim:al_pdf,rieck:dynamic,schultz:data_mining,shabtai:survey} provide good examples of static and dynamic information to extract. So far only the static imports made by each executable were extracted (i.e. \gls{dll} functions), leaving much information to be extracted still. To speed parsing, Celery\cite{tool:celery} was used to distribute the work. Since it has proven to be useful in extracting static imports, it should also be applied for the remaining parsing.

\begin{figure}[!h]
	\subfloat[Total samples per month]{
		\includegraphics[width=\textwidth]{samples_count}
	}\\
	\subfloat[Percentage of duplicated (same MD5) samples per month]{
		\includegraphics[width=\textwidth]{samples_dups}
	}
	\caption{Statistics for the scraped submissions. Count and percentage of duplicates}
	\label{fig:dist_submissions}
\end{figure}

\begin{figure}[!h]
	\includegraphics[width=\textwidth]{samples_types}
	\caption{Statistics for the scraped submissions. Distribution of file types}
	\label{fig:dist_submissions_types}
\end{figure}

\subsection{Data Labeling}\label{subsec:data_labeling}

While the previous section dealt with extracting information about an executable, the current section deals with labeling each executable as malicious or benign. This task is not trivial, as each sample is classified with different labels by different vendors. Not only that, but the number of anti-virus classifications varies among samples, as only vendors that have seen the sample classify it. Metrics have to be defined as to split the samples into malware or goodware.

To help labeling the dataset, this work proposes to filter samples by only keeping samples classified by a subset of vendors. The chosen subset is created by ordering the vendors by the number of classified samples (from the total number of samples) and choosing the top 20. This guarantees that every vendor has a classification for the sample at a reduced number of total samples.

To label a sample as benign, the current work defines that if a sample is considered as clean by all vendors, it is labeled as goodware. When dealing with duplicated samples, the most recent report is taken into account. Although this method may seen obvious in the first place, it can fail if at the time the report was made, vendors weren't able to detect the threat and labeled the sample benign. As authors in \cite{miller:rev_int} have shown, this problem occurred in almost 30\% of their duplicated samples.

To label a sample as malicious, the number of detections has to be taken into account. The imposed question is at which threshold should a sample be considered malware? On top of that, the type of malware should also be taken into consideration to provide a uniform (i.e balanced) dataset, as mentioned in \cite{rossow:practices}. A possible method to help balance the malware variety is to adopt some of the work in \cite{perdisci:behavior} regarding the naming problem: creating a graph where the nodes are all existing labels and edges connect each node based on how many times labels appear together in a sample.

Regarding data labeling some work was done to better understand the problem. For the top 20 anti-virus vendors, a filter was applied as to limit samples where all vendors classified as either goodware or malware, resulting in a total of 269,563 samples. A histogram of how many vendors classified a sample as malware was created, resulting in figure \ref{fig:dist_mal_class}. This shows how the threshold to label a sample as malware can be a difficult choice, since vendors often disagree on what is malware.

\begin{figure}[]
	\includegraphics[width=\textwidth]{samples_top20_class}
	\caption{Distribution of how many vendors classify a sample as malware.}
	\label{fig:dist_mal_class}
\end{figure}

Further work was done to understand the naming variation. Again for the top 20 anti-virus vendors a filter was applied as to limit samples where all vendors classified as malware, resulting in a total of 10,868 (as can be seen in the last column of figure \ref{fig:dist_mal_class}). The unique number of names was counted, together with the percentage of repeating names. The results can be seen in table \ref{tab:statistics}, showing a varying number of unique names per vendor, with an average of 2,403 unique malware names per vendor. For the ten most repeating names, it can be seen that the variation is smaller and that on average the top ten most repeating names account for 28.7\% of the samples.

\subsection{Classifier Implementation}\label{subsec:class_implementation}

This final stage takes the work done in the previous stages to generate a model that is able to detect malware based on static and dynamic features. This stage is to take into consideration the guidelines given in \cite{rossow:practices,shabtai:survey} and the classifiers in \cite{miller:rev_int,nissim:al_pdf,rieck:dynamic,schultz:data_mining}. Ideally different classifiers should be tested, as some perform better over others for some set of features. The possibility of using multiple classifiers for different features can also be applied to improve the overall result.

\paragraph{Evaluation.} The final result is to be measured using the common evaluation metrics that others\cite{miller:rev_int,nissim:al_pdf,schultz:data_mining} have applied, specifically the \gls{tpr} and \gls{fpr}. The \gls{tpr} measures the detection rate of the resulting model, while the \gls{fpr} measures the fall-out.

The described metrics are to be applied in two different conditions: laboratory and real-world. For laboratory conditions, the dataset is to be randomly split, using one part for training and other for evaluation. For real-world conditions, the dataset is to be split by date, using older samples for training and more recent samples for evaluation. By forcing this temporal consistency, the current work will also try to validate the assumption in \cite{miller:rev_int}.

\subsection{Experimental Work}\label{subsec:exp_work}

As a motivation for the work still to be done, a small experiment was conducted to check how easily the creation of a classifier to detect malware is. A dataset was created using the static imports as features. As for the sample labeling, the classification as given by Microsoft's Windows Defender was used.

Using scikit-learn\cite{tool:sklearn} each sample was converted to a binary vector, where each position corresponded to the presence (or absence) of a static import. This feature representation lead to vectors with nearly 30,000 positions (i.e. there were almost 30,000 different static imports). A logistic regression classifier was trained with 75\% of the samples, using the remaining 25\% for evaluation. To check how the dataset balancing affected the classifier, as mentioned in \cite{rossow:practices}, three variants of the dataset were used.

The first dataset was the default and contained more goodware than malware (154,556 goodware, 76,495 malware), this yielded a detection rate 57.2\% with a \gls{fpr} of 28.4\%. The second dataset used the same amount of goodware and malware (76,495 samples each), resulting in a detection rate of 69.2\% and a \gls{fpr} of 23.7\%. The last dataset contained more malware than goodware (76,495 malware, 38,247 goodware) and gave a detection rate of 92\% with a \gls{fpr} of 22.4\%. Table \ref{tab:experimental} summarizes the results and notes the results from \cite{miller:rev_int} as a comparison.

\begin{table}[h!]
	\centering
	\def\arraystretch{1.5}
	\begin{tabular}{lcc}
		\toprule
		& Detection Rate        & False Positive Rate        \\
		\midrule
		First Dataset  & 0.572     & 0.284      \\ \hline
		Second Dataset & 0.692     & 0.237      \\ \hline
		Third Dataset  & 0.92      & 0.224      \\ \hline
		Miller et al   & [0.65, 0.75] & [0.001, 0.01] \\
		\bottomrule
	\end{tabular}
	\caption{Summary of experimental results compared to Miller et al\cite{miller:rev_int}}
	\label{tab:experimental}
\end{table}

This quick test gave some interesting results, although they do not compare to related work, it showed that with little to no work on the dataset and classifier a simple model was achieved. It also provided some experience with scikit-learn, which is to be used in the work to come.\\

The work that will result in the proposed system is to be developed between June 2017 and March 2018.

The planned schedule is to use the month of June to finish the data collection phase, together with the labeling phase during the months of June and July. The final phase of implementation is planned to take six months, from August to January, as many variables are present and more experiments should provide better results. Regarding the writing of the thesis document, the months of February and March are to be used. A more detailed planning can be seen in figure \ref{fig:schedule}.

\begin{figure}[h]
	\includegraphics[width=\textwidth]{gantt}
	\caption{Planned Schedule}
	\label{fig:schedule}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%CONCLUSION%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}\label{sec:conclusion}

With the growing amount of malware instances, being able to effectively detect malware as it appears is a must, otherwise cases like WannaCry malware\footnote{Wanna Cry Profile [https://www.fireeye.com/blog/threat-research/2017/05/wannacry-malware-profile.html]} will continue to have real and serious effects\footnote{Huffing Post Article [http://www.huffingtonpost.com/entry/britain-cyber-attack-national-health-service\_us\_5915e62ae4b00f308cf4fc86]}.

When new types of malware arise, anti-virus may be unable to detect them with the usual signature and heuristic methods. The application of \gls{ml} to the detection task as shown to be possible and effective\cite{kolter:learning,miller:rev_int,nissim:al_pdf,perdisci:behavior,rieck:dynamic,schultz:data_mining}. With that in mind, the current work will develop a \gls{ml} system capable of detecting malicious files, as well as test how different types of evaluations can impact the effectiveness of the solution.

The proposed solution, as described in section \ref{sec:solution}, is to be split in three steps: data collection, data labeling and classifier implementation. Some work as been done in each step as to get a feel on what to expect and the problems that might surge along the way.

The next step is to fully terminate each of the sequential stages, with special focus on classifier implementation as many variations can be tested, providing a better research and final result.

Hopefully the current work enlightened the reader on the problem, how the current technologies deal with it and how the proposed work could improve the detection task.

\begin{landscape}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                      %
%     File: Thesis_Background.tex                                      %
%     Tex Master: Thesis.tex                                           %
%                                                                      %
%     Author: João C. Godinho                                          %
%     Last modified : Apr 2018                                         %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Related Work}
\label{chapter:related_work}

The following chapter will go over prior work that closely relates to the topic of our research and areas of contribution.
We approach the definition of malware and how it can hard to detect and label by providing some background on the subject.
We then proceed to elucidate on different malware detection techniques, from simple methods to more complex approaches, focusing on a specific malware analysis technique that provided the basis for our work.

We advance to the general topic of \gls{ml}, after which we describe related work that studies the use of \gls{ml} in malware detection.
Lastly, we reference prior work that not only tackles our same goal, but also takes into account the impact of laboratory \textit{vs.} real world scenarios, as well as the notion of temporal based detection, or concept drift.

\newpage
\todo[inline]{CSF Related work}

We use this section to present prior work that closely relates to the topic of our research and our areas of contribution. Specifically, detecting malicious software by training supervised models on static information, and validation methodologies that resemble real-world conditions.

Shabtai et al.\cite{shabtai:survey} provide a survey directed
at the application of \gls{ml} classifiers to detect malware from static
features. Their work concerns the design and evaluation of such systems. Our contributions are inspired by the concern of how to correctly evaluate \gls{ml} classifiers, regarding size, reliability of labeling metrics and chronological evaluation.

In the topic of methodologies that resemble real-world conditions, Srndi\'c et al.\cite{vsrndic2013detection}
train and validate their malicious PDF detector under laboratory and real-world conditions. Laboratory conditions consist on applying regular cross-validation, whereas real-world conditions validate a newer dataset with a model created from outdated data (i.e.\ older then the validation), and also validate the model when the validation set spans one week and the training is gathered in the previous 4 weeks. They show that laboratory conditions inflate the results, when compared to real-world conditions.

Miller et al.\cite{miller:rev_int} also introduce sample temporal consistency. They show the impact of performance measurement technique on a dataset containing 1.1 million samples, when using cross-validation and temporally consistent samples. As noted by others, regular cross-validation showed inflated results when compared to temporally consistent samples.

Our work enhances these methodologies by analyzing the performance variation when the distance between the training and validation set increases and decreases, as well as analysis on how reducing the size of the training without compromising the results.

Kolter et al.\cite{kolter:learning} learn to detect malicious executables in a dataset with under 4,000 samples, obtained from reliable sources, and evaluate their model under standard cross-validation and by gathering newer malware samples to validate for new and unseen samples. They show how the model provides optimal results under cross-validation, but for the unseen samples lower scores are obtained. Our work considers these results to compare how reliability affects performance.

Sebastián, M. et al.\cite{sebastian2016avclass} develop \textit{AVClass}, a tool that given a set of antivirus vendors, outputs the most likely family name. They test their tool under 10 datasets, totaling 8.9 M samples, with results showing an F1 measure up to 93.9 on labeled datasets. The tool takes as input the labels as seen in VirusTotal, tokenizes the labels, replaces known aliases and general names (e.g.\ win32, trojan, generic), ending with possible family names. These remaining names are counted and the most frequent one is given as family name. Our work takes advantage of this tool, not to label malware families, but minimal modifications, to label more general malware classes (e.g.\ trojan, virus).

Deo, A. et al.~\cite{deo2016prescience} focus over the problem of \gls{ml} models becoming antiquated over time, given how malware evolves. This problem is seen as concept drift, where the performance of a model over time diminishes, as the statistical properties of malware (i.e.\ features) change over time. They study how probabilistic predictors can help minimize the aforementioned problem, by indicating when retraining of a given model is necessary.

Jordaney, R. et al.~\cite{jordaney2017transcend} also study the problem of concept drift in malware classification models, focusing on providing metrics, based on statistical comparison of samples, to detect when should a model be retrained.

Both ~\cite{deo2016prescience} and ~\cite{jordaney2017transcend} are on the subject of our work, regarding the problem of concept drift, but differ on the study-case. In our work we acknowledge the concept but focus on its relative effects and how one can balance the size of needed training data vs. validation data, when maintaining a temporal consistent dataset. Whereas their work provide indicators for when should a model be retrained when the problem of concept drift becomes significant.

\newpage
\todo[inline]{Project Report related work}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}
\label{section:background}

Malware, which can be referred to by other names like malicious software or malicious code, can be described as \say{a program that has malicious intent}\cite{christodorescu:semantics}, or more detailed as any code added, changed, or removed from a software system in order to intentionally cause harm or subvert the intended function of the system\cite{mcgraw:mal_code}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Malware Detection}
\label{section:mal_detec}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Cuckoo Sandbox and Malwr}
\label{section:cuckoo}

@@

%\begin{figure}[h]
%	\includegraphics[width=\textwidth]{malwr_sample}
%	\caption{Example of a behavioral report from Malwr}
%	\label{fig:malwr_sample}
%\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Machine Learning}
\label{section:ml}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Machine Learning in Malware Detection}
\label{section:ml_md}

@@

Before analyzing the work done with \gls{ml} in malware detection, it is worth noting how such work is to be conducted. Rossow et al\ \cite{rossow:practices} present some guidelines for designing malware experiments. Their work surveys 36 academic publications from 2006 to 2011 that rely on malware execution. The authors note frequent shortcomings in the surveyed publications regarding assumptions on the use of execution-driven datasets, absence of security precautions taken during experiments and insufficient description of the experimental setup. The provided guidelines are split into four groups: correct datasets, transparency, realism and safety.

@@

%\begin{figure}[h]
%	\includegraphics[width=\textwidth]{mist}
%	\caption{\gls{mist} representation as seen in \cite{rieck:dynamic}}
%	\label{fig:mist}
%\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Theoretical Overview}
%\label{section:overview}
%
%Some overview of the underlying theory about the topic...
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Theoretical Model 1}
%\label{section:theory1}
%
%The research should be supported with a comprehensive list of references.
%These should appear whenever necessary, in the limit, from the first to the last chapter.
%
%A reference can be cited in any of the following ways:
%%
%\begin{itemize}
%  \item Citation mode \#1 - \quad \cite{jameson:adjointns}
%  \item Citation mode \#2 - \quad \citet{jameson:adjointns}
%  \item Citation mode \#3 - \quad \citep{jameson:adjointns}
%  \item Citation mode \#4 - \quad \citet*{jameson:adjointns}
%  \item Citation mode \#5 - \quad \citep*{jameson:adjointns}
%  \item Citation mode \#6 - \quad \citealt{jameson:adjointns}
%  \item Citation mode \#7 - \quad \citealp{jameson:adjointns}
%  \item Citation mode \#8 - \quad \citeauthor{jameson:adjointns}
%  \item Citation mode \#9 - \quad \citeyear{jameson:adjointns}
%  \item Citation mode \#10 - \quad \citeyearpar{jameson:adjointns}
%\end{itemize}
%%
%Several citations can be made simultaneously as \citep{nocedal:opt,marta:ijcfd}. \\
%
%This is often the default bibliography style adopted (numbers following the citation order), according to the options:\\
%{\tt \textbackslash usepackage\{natbib\}} in file {\tt Thesis\_Preamble.tex},\\
%{\tt \textbackslash bibliographystyle\{abbrvnat\}} in file {\tt Thesis.tex}.\\
%
%Notice however that this style can be changed from numerical citation order to authors' last name with the options: \\
%{\tt \textbackslash usepackage[numbers]\{natbib\}} in file {\tt Thesis\_Preamble.tex},\\
%{\tt \textbackslash bibliographystyle\{abbrvunsrtnat\}} in file {\tt Thesis.tex}. \\
%
%Multiple citations are compressed when using the {\tt sort\&compress} option when loading the {\tt natbib} package as {\tt \textbackslash usepackage[numbers,sort\&compress]\{natbib\}} in file {\tt Thesis\_Preamble.tex}, resulting in citations like \citep{marta:ijcfd1,marta:ijcfd2,marta:ijcfd3,marta:ijcfd4}.
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Theoretical Model 2}
%\label{section:theory2}
%
%Other models...
